---
title: "CTDquerier: Case study on Environmental Chemicals and Human Non-Comunicable Diseases"
author:
- name: Carles Hernandez-Ferrer
  affiliation: ISGlobal, Centre for Research in Environmental Epidemiology (CREAL)
- name: Juan R. Gonzalez
  affiliation: ISGlobal, Centre for Research in Environmental Epidemiology (CREAL)
  email: juanr.gonzalez@isglobal.org
date: "`r doc_date()`"
package: "`r pkg_ver('CTDquerier')`"
bibliography: case_study.bib
csl: biomed-central.csl
vignette: >
  %\VignetteIndexEntry{CTDquerier: Case study on Environmental Chemicals and Human Non-Comunicable Diseases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::html_document2:
    toc_float: true
---

# Introduction

## Health Impacts of Chemicals

The health impact of chemicals is determined by a process of assessment which aims to provide a consensus scientific description of the risks of chemical exposures. The International Programme on Chemical Safety (IPCS) of the World Health Organization (WHO) is one of the organizations that works to establish the scientific basis for the sound management of chemicals, and to strengthen national capabilities and capacities for chemical safety. In its responsibilities are the publication of the reports and other related documents so that governments and international and national organizations can use them as the basis for taking preventive actions against adverse health and environmental impacts.

The leading cause of deaths and disability worldwide is ischaemic heart  disease. The second largest contributor to global mortality is stroke. The 35% of the first and the 42% of the second could be prevented by reducing or removing exposure to chemicals such as from  ambient air pollution, household air pollution, second-hand smoke and lead [@WHOHealthEnvironment2016].

Chemicals such as heavy metals, pesticides, solvents, paints, detergents, kerosene, carbon monoxide and drugs lead to unintentional poisonings at home and in workplace. This type of unintentional poisonings are estimated to cause 193 000 deaths annually [@WHOHealthEnvironment2016][@WHOKnownsUnknowns2016].

It is well known that air pollution and second-hand smoke are risk factors for adverse pregnancy outcomes like low birth weight, prematurity and stillbirths [@Khader2011]. Antenatal exposure to second-hand smoke for example was estimated to increase the overall risk for stillbirths by 23% and for congenital malformations by 13%. There are, furthermore, potential links between various chemicals and adverse pregnancy outcomes or congenital malformations, though evidence is limited [@WHOHealthEnvironment2016].

Moreover, second-hand smoke and air pollution can lead to the development of asthma. Air pollution additionally provokes asthma exacerbations and increases related hospital admissions [@WHOHealthEnvironment2016][@Macintyre2013].


## The Comparative Toxicogenomics Database

The Comparative Toxicogenomics Database (*CTDbase*; http://ctdbase.org) is a public resource for toxicogenomic information manually curated from the peer-reviewed scientific literature, providing key information about the interactions of environmental chemicals with gene products and their effect on human disease [@CTDbase2003][@CTDbase2017].

*CTDbase* is curated by professional biocurators who leverage controlled vocabularies, ontologies and structured notation to code a triad of core interactions describing chemical-gene, chemical-disease and gene-disease relationships, which are then internally integrated to generate inferred chemical-gene-disease networks [@CTDbase2017][@CTDbase2011].

*CTDbase* if offered to public by using a web-baed inetrface that includes basic and advanced query options to access data for sequences, references, and toxic agents, and a platform for analyzing sequences.

*CTDbase* includes more than 30.5 million toxicogenomic connections relating chemicals/drugs, genes/proteins, diseases, exposures, Gene Ontology (GO) annotations, pathways (KEGG/Reactome), and gene interaction modules [@CTDbase2017].


## `CTDquerier` R package

`CTDquerier` is an R package that allows to R users to download basic data from *CTDbase* about genes, chemicals and diseases. Once the user's input is validated allows to query *CTDbase* to dowload the information of the given input from the other modules. Tables \@ref(tab:CTDquerier-queries) indicates the information `CTDquerier` can download from *CTDbase* depending on the user's input.

Input Type | Gene Interaction | Disease | Pathway | Gene Ontology | Chemical Interaction |
:----------|:----------------:|:-------:|:-------:|:-------------:|:--------------------:|
Genes      | X                | X       | X       | X             | X                    |
Chemical   | X                | X       | X       | X             |                      |
Disease    | X                |         | X       |               | X                    |

: (\#tab:CTDquerier-queries) Relation between input accepted by `CTDquerier` and the data it is capable to download from *CTDbase*.



### Installation

## Objectives


# Getting Started



Case-control admixture mapping in Latino populations enriches for known asthma-associated genes 
http://www.sciencedirect.com/science/article/pii/S0091674912003703

Maternal smoking in pregnancy and its influence on childhood asthma
https://www-ncbi-nlm-nih-gov.sare.upf.edu/pmc/articles/PMC5034599/

Genome-wide profiling identifies epithelial cell genes associated with asthma and with treatment response to corticosteroids
http://www.pnas.org/content/104/40/15858.full





The Effects of Environmental Toxins on Allergic Inflammation
https://www-ncbi-nlm-nih-gov.sare.upf.edu/pmc/articles/PMC4214967/


```{r show_statistics}
plot(1:4, 1:4)
```





# Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

# Bibliography
