---
title: "CTDquerier: A package for downstream analysis and data visualization with CTDbase data"
author:
- name: Carles Hernandez-Ferrer
  affiliation: ISGlobal, Centre for Research in Environmental Epidemiology ( CREAL )
- name: Juan R. Gonzalez
  affiliation: ISGlobal, Centre for Research in Environmental Epidemiology ( CREAL )
  email: juanr.gonzalez@isglobal.org
date: "`r doc_date()`"
package: "`r pkg_ver( 'CTDquerier' )`"
csl: biomed-central.csl
bibliography: case_study.bib
vignette: >
    %\VignetteIndexEntry{CTDquerier: A package for downstream analysis and data visualization with CTDbase data}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
output: 
    BiocStyle::html_document:
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

## The Comparative Toxicogenomics Database

The Comparative Toxicogenomics Database (*CTDbase*; http://ctdbase.org) is a public resource for toxicogenomic information manually curated from the peer-reviewed scientific literature, providing key information about the interactions of environmental chemicals with gene products and their effect on human disease [@CTDbase2003][@CTDbase2017]. *CTDbase* is offered to public by using a web-based interface that includes basic and advanced query options to access data for sequences, references, and toxic agents, and a platform for analysing sequences.

## `CTDquerier` R package

`CTDquerier` is an R package that allows to R users to incorporate basic data from *CTDbase* about genes, chemicals and diseases into their analyses. First it loads *CTDbase* files into R, then it allows to validate user's input and then to extract relationships between the diferent sets of data.

`CTDquerier` can be installed using `devtools`. To install `CTDquerier` run the following command in an R session:

```{r, eval=FALSE}
devtools::install_github(
  "https://github.com/isglobal-brge/CTDquerier", 
  repo = "devel"
)
```

Once installed, `CTDquerier` should be loaded running the following command:

```{r}
library(CTDQuerier)
```


The classes from `CTDquerier` can be classified in three groups according to their "function" in the _data-workflow_:

1. Data loading:

| Input     | Class           | Description                        |
|:----------|:----------------|:-----------------------------------|
| Genes     | `ctdVocabulary` | It loads vocabulary files.         |
| Diseases  | `ctdVocabulary` | It loads vocabulary files.         |
| Chemicals | `ctdVocabulary` | It loads vocabulary files.         |
| Pathways  | `ctdVocabulary` | It loads vocabulary files.         |
| `list`    | `ctdStorage`    | It loads database (storage) files. |

2. Data querying:

| Input     | Class           | Method/Function | Description                                                           |
|:----------|:----------------|:----------------|:----------------------------------------------------------------------|
| Genes     | `ctdVocabulary` | `search_term`   | Searches a single or multiple gene terms from a specific source.      |
| Diseases  | `ctdVocabulary` | `search_term`   | Searches a single or multiple disease terms from a specific source.   |
| Chemicals | `ctdVocabulary` | `search_term`   | Searches a single or multiple chemical terms from a specific source.  |
| Pathways  | `ctdVocabulary` | `search_term`   | Searches a single or multiple pathway terms from a specific source.   |
| Genes     | `ctdVocabulary` | `validate_term` | Validates a single or multiple gene terms from a specific source.     |
| Diseases  | `ctdVocabulary` | `validate_term` | Validates a single or multiple disease terms from a specific source.  |
| Chemicals | `ctdVocabulary` | `validate_term` | Validates a single or multiple chemical terms from a specific source. |
| Pathways  | `ctdVocabulary` | `validate_term` | Validates a single or multiple pathway terms from a specific source.  |
|   | `...` | It loads and stored network files. |

3. Data visualization:

| Input     | Class           | Method          | Description                                                  |
|:----------|:----------------|:----------------|:-------------------------------------------------------------|
| ... | `...` | `...` | ... |


# Using *CTDbase* in R

## Loading vocabulary

```{r, echo=FALSE}
ctd_file <- function(file) {
  return(paste0("/Users/chernandez/Downloads/ctdbase/", file))
}
```

The first step to use `CTDquerier` is to load the vocabulary files. 

To do so, download them from [http://ctdbase.org/downloads/](http://ctdbase.org/downloads/). They are tagged as _[13. Chemical vocabulary](http://ctdbase.org/downloads/#allchems)_, _[14. Disease vocabulary (MEDIC)](http://ctdbase.org/downloads/#alldiseases)_, _[15. Gene vocabulary](http://ctdbase.org/downloads/#allgenes)_, and _[16. Pathway vocabulary](http://ctdbase.org/downloads/#allpathways)_. Once on the disk, they are loaded using the class `ctdVocabulary`:

```{r}
genVocabulary <- ctdVocabulary$new(ctd_file("CTD_genes.csv.gz"), "gene")
disVocabulary <- ctdVocabulary$new(ctd_file("CTD_diseases.csv.gz"), "disease")
chemVocabulary <- ctdVocabulary$new(ctd_file("CTD_chemicals.csv.gz"), "chemical")
pathVocabulary <- ctdVocabulary$new(ctd_file("CTD_pathways.csv.gz"), "pathway")
```

The previous piece of code showed the function `ctd_file`. This is auxiliary function that creates a path to a given file and it will be used along the _vignnette_ in order to make the paths shorter. It's implementation is:

```{r, eval=FALSE}
ctd_file <- function(file) {
  return(paste0("/path/to/ctdbase/files/", file))
}
```

## Loading storage

In order to perform queries to _CTDbase_, `CTDquerier` needs the _"database files"_. Download them from [http://ctdbase.org/downloads/](http://ctdbase.org/downloads/). They are:

 1. Chemical–gene interactions
 2. Chemical–gene interaction types
 3. Chemical–disease associations
 4. Chemical–GO enriched associations
 5. Chemical–pathway enriched associations
 6. Gene–disease associations
 7. Gene–pathway associations
 8. Disease–pathway associations
 9. Chemical–phenotype interactions

These files are given to `CTDquerier` as `list` of vectors, pointing to their location in disk:

```{r}
fileList <- list(
  c("Chemical–gene interactions", ctd_file("CTD_chem_gene_ixns.csv.gz")),
  c("Chemical–gene interaction types", ctd_file("CTD_chem_gene_ixn_types.csv")),
  c("Chemical–disease associations", ctd_file("CTD_chemicals_diseases.csv.gz")),
  c("Chemical–GO enriched associations", ctd_file("CTD_chem_go_enriched.csv.gz")),
  c("Chemical–pathway enriched associations", ctd_file("CTD_chem_pathways_enriched.csv.gz")),
  c("Gene–disease associations", ctd_file("CTD_genes_diseases.csv.gz")),
  c("Gene–pathway associations", ctd_file("CTD_genes_pathways.csv.gz")),
  c("Disease–pathway associations", ctd_file("CTD_diseases_pathways.csv.gz")),
  c("Chemical–phenotype interactions", ctd_file("CTD_pheno_term_ixns.csv.gz"))
)
```

Having the list of files, the method `load_storage` of `ctdStorage` will load them and prepare them to our use.

```{r}
fileStorage <- ctdStorage$new(fileList)
```

## Searching and validatign terms

With the vocabulary in the R session we can now validate the terms to query _CTDbase_. The function `search_term` provides with a searching tool when we don't know if the terms we want to query are in  _CTDbase_. This function returns a table with the given terms and all the terms in a `ctdVocabulary` object, joint with a distance measurement.

For instance, we want to find which is the right terminology for "Autism" (are general spectrum disorder) in _CTDbase_:

```{r}
tmp <- disVocabulary$search_term("autism spectrum")
head(tmp)
```
Another case is to look for environmental contaminants like "air pollution", that in _CTDbase_ appears as "air pollutants":

```{r}
tmp <- chemVocabulary$search_term("air pollution")
head(tmp)
```
Once we know the right term(s) to use we validate them:

```{r, eval=FALSE}
geneValid <- genVocabulary$validate_term(
  terms = c( "APOE", "APOEB", "APOE2", "APOE3" , "APOE4", "APOA1", "APOA5" )
)
geneValid
```

The method `get_terms` (from any `ctdTerm` object) allows to list the valid and invalid terms:

```{r ctd_get_genes}
geneValid$get_terms()
```





